Optimal strategies in mental rotation tasks

### 2013-05-13 man/final-project.tex
Move shepard rotation figure
Comment out abstract and acknowlegements
Spellcheck

\begin{abstract}
\TODO{}
\end{abstract}

@@

In the case of the classic mental rotation task, we might ask: in what
direction should the object be rotated?  What are the requirements for
``congruency''? When should one stop rotating and accept the
hypothesis that the objects are different? In this project, I will
investigate the optimal computational solution to this task. Thus, the
specific question this project aims to answer is: what is the rational
computational solution to the mental rotation task, and does it
predict the people's behavior on the task?

@@

A more flexible strategy uses what is known as \emph{Bayesian
  Quadrature} \cite{Diaconis:1988uo,OHagan:1991tx} to estimate $Z$.
Bayesian Quadrature allows us to compute a posterior distrbution over
$Z$ by placing a Gaussian Process (GP) prior on the function $S$ and
evaluating $S$ at a particular set of points. Because our data is
circular, we can use a periodic kernel \cite{Rasmussen:2006vz}:
\begin{equation}
k(R, R^\prime)=h^2\exp\left(-\frac{2\sin^2\left(\frac{1}{2}(R-R^\prime)\right)}{w^2}\right)
\end{equation}

Bayesian Quadrature has its difficulties, however. While in our case
$S$ is a non-negative likelihood function, GP regression enforces no
such constraint. In an effort to avoid this problem,
\cite{Osborne:2012tm} give a method which involves instead placing a
prior over the log likelihood\footnote{In practice, as in
  \cite{Osborne:2012tm}, we use the transform of
  $\log(S+1)$. Additionally, while \cite{Osborne:2012tm} use a
  combination of MLII and marginalization to fit the kernel
  parameters, we set the output scale $h_S=0.125$ and $h_{\log
    S}=\log(h_S + 1)$, and use MLII to fit all other parameters.},
thus ensuring that $S=e^{\log S}$ will be positive\footnote{We are not
  guaranteed positivity, however, the approximation to the integral
  over $\log S$ (Equation \ref{eq:bq-Z-mean}) requires computing
  $\mu_S$, which may have nonpositive segments.}:
\begin{equation*}
  E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\ \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S}, \Sigma_{\log S}\right)\ \mathrm{d}\log S
\end{equation*}
where $\mu_{\log S}$ and $\Sigma_{\log S}$ are the mean and
covariance, respectively, of the GP regression over $\log S$ given
$\mathbf{R}$. We approximate this according to the method given in
\cite{Osborne:2012tm}:
\begin{equation}
  \mu_Z=E[Z\ \vert \ S, \log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\ \mathrm{d}R 
  \label{eq:bq-Z-mean}
\end{equation}
where $\mu_S$ is the mean of a GP regression over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ is a regression over
$\Delta_c=\mu_{\log S} - \log \mu_S$ given $\mathbf{R}_c$, which
consists of $\mathbf{R}$ and a set of intermediate \emph{candidate
  points}\footnote{These candidate points do not require evaluating
  the true $S$, only the GP estimates of $S$ and $\log S$.} $c$ as
described in \cite{Osborne:2012tm}. The variance of the estimate of
$Z$ is given by:
\begin{equation}
  \sigma_Z=\mathrm{Var}(Z\ \vert \ S, \log S, \Delta_c) = \int_R\int_{R^\prime} \mu_S(R)\mu_S(R^\prime) \Sigma_{\log S}(R, R^\prime)p(R)p(R^\prime)\ \mathrm{d}R\ \mathrm{d}R^\prime
  \label{eq:bq-Z-var}
\end{equation}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Each subplot shows the
    correspondence between the true angle of rotation ($R$) for $h_1$
    stimuli and the amount of rotation performed by the model. Dotted
    lines indicate a perfect 1:1 relationship. The \naive{} and
    parametric models (left and center panels, respectively) are
    identical; they show modest agreement until rougly
    $\frac{\pi}{2}$, after which agreement decreases as they tend to
    encounter local maxima. The nonparametric model (right panel)
    exhibits a significant correlation between the true angle and the
    amount the model rotates.}
  \label{fig:rotations}
\end{figure}

@@

The correlation between the number of rotations computed by the
\naive{} model and the true angle of rotation was $\rho=-0.11$. The
shape of the data is more interesting (see Figure \ref{fig:rotations},
left panel): the \naive{} model actually corresponds quite well to
the true angle of rotation for $R<\frac{\pi}{2}$ ($\rho=0.64$). This
is unsurprising, because the closer the true angle is to zero, the
less the model has to rotate, and the less likely it will get stuck on
local maxima. Thus, it is more likely to locat the global maximum,
which generally corresponds to the true angle of rotation. For
$R>\frac{\pi}{2}$, we see an increasing tendency to under-rotate
($\rho=-0.42$); this is likely because it finds a local maxima and
ends prematurely.

@@

Similarly, because the nonparametric model actively attemps to obtain
an accurate estimate of $Z$, its estimate of $Z$ is indeed fairly
accurate ($\MSE{}=0.10$). This is slightly higher than the
nonparametric model, but is due to a few outliers. If we again exclude
stimuli for which $Z$ is greatly overestimated ($Z\geq 0.3$), the
nonparametric model's error decreases significantly ($\MSE{}=0.03$).

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/Z_accuracy.pdf}
  \caption{\textbf{Accuracy in estimating $Z$.} Each subplot shows the
    true (``gold standard'') value of $Z$ vs. the value estimated by
    the model. Black dotted lines indicate a perfect 1:1
    correspondence between the true and estimated values. All models
    perform well for $h_0$ stimuli, but exhibit different behavior for
    $h_1$ stimuli. The \naive{} model (left panel) tends to either
    greatly overestimate, or underestimate. The parametric model
    (center panel) is very accurate for about half the $h_1$ stimuli,
    and severly underestimates the rest. The nonparametric model
    (right panel) maintains a decent correspondence with the true $Z$,
    with the exception of a handful of outliers.}
  \label{fig:accuracy}
\end{figure}

@@

\textbf{Acknowledgements}

\TODO{}

>>>>

% \begin{abstract}
% \TODO{}
% \end{abstract}

@@

In the case of the classic mental rotation task, we might ask: in what
direction should the object be rotated?  What are the requirements for
``congruence''? When should one stop rotating and accept the
hypothesis that the objects are different? In this project, I will
investigate the optimal computational solution to this task. Thus, the
specific question this project aims to answer is: what is the rational
computational solution to the mental rotation task, and does it
predict the people's behavior on the task?

@@

A more flexible strategy uses what is known as \emph{Bayesian
  Quadrature} \cite{Diaconis:1988uo,OHagan:1991tx} to estimate $Z$.
Bayesian Quadrature allows us to compute a posterior distribution over
$Z$ by placing a Gaussian Process (GP) prior on the function $S$ and
evaluating $S$ at a particular set of points. Because our data is
circular, we can use a periodic kernel \cite{Rasmussen:2006vz}:
\begin{equation}
k(R, R^\prime)=h^2\exp\left(-\frac{2\sin^2\left(\frac{1}{2}(R-R^\prime)\right)}{w^2}\right)
\end{equation}

Bayesian Quadrature has its difficulties, however. While in our case
$S$ is a non-negative likelihood function, GP regression enforces no
such constraint. In an effort to avoid this problem,
\cite{Osborne:2012tm} give a method which involves instead placing a
prior over the log likelihood\footnote{In practice, as in
  \cite{Osborne:2012tm}, we use the transform of
  $\log(S+1)$. Additionally, while \cite{Osborne:2012tm} use a
  combination of MLII and marginalization to fit the kernel
  parameters, we set the output scale $h_S=0.125$ and $h_{\log
    S}=\log(h_S + 1)$, and use MLII to fit all other parameters.},
thus ensuring that $S=e^{\log S}$ will be positive\footnote{We are not
  guaranteed positivity, however, the approximation to the integral
  over $\log S$ (Equation \ref{eq:bq-Z-mean}) requires computing
  $\mu_S$, which may have non-positive segments.}:
\begin{equation*}
  E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\ \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S}, \Sigma_{\log S}\right)\ \mathrm{d}\log S
\end{equation*}
where $\mu_{\log S}$ and $\Sigma_{\log S}$ are the mean and
covariance, respectively, of the GP regression over $\log S$ given
$\mathbf{R}$. We approximate this according to the method given in
\cite{Osborne:2012tm}:
\begin{equation}
  \mu_Z=E[Z\ \vert \ S, \log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\ \mathrm{d}R 
  \label{eq:bq-Z-mean}
\end{equation}
where $\mu_S$ is the mean of a GP regression over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ is a regression over
$\Delta_c=\mu_{\log S} - \log \mu_S$ given $\mathbf{R}_c$, which
consists of $\mathbf{R}$ and a set of intermediate \emph{candidate
  points}\footnote{These candidate points do not require evaluating
  the true $S$, only the GP estimates of $S$ and $\log S$.} $c$ as
described in \cite{Osborne:2012tm}. The variance of the estimate of
$Z$ is given by:
\begin{equation}
  \sigma_Z=\mathrm{Var}(Z\ \vert \ S, \log S, \Delta_c) = \int_R\int_{R^\prime} \mu_S(R)\mu_S(R^\prime) \Sigma_{\log S}(R, R^\prime)p(R)p(R^\prime)\ \mathrm{d}R\ \mathrm{d}R^\prime
  \label{eq:bq-Z-var}
\end{equation}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Each subplot shows the
    correspondence between the true angle of rotation ($R$) for $h_1$
    stimuli and the amount of rotation performed by the model. Dotted
    lines indicate a perfect 1:1 relationship. The \naive{} and
    parametric models (left and center panels, respectively) are
    identical; they show modest agreement until roughly
    $\frac{\pi}{2}$, after which agreement decreases as they tend to
    encounter local maxima. The nonparametric model (right panel)
    exhibits a significant correlation between the true angle and the
    amount the model rotates.}
  \label{fig:rotations}
\end{figure}

@@

The correlation between the number of rotations computed by the
\naive{} model and the true angle of rotation was $\rho=-0.11$. The
shape of the data is more interesting (see Figure \ref{fig:rotations},
left panel): the \naive{} model actually corresponds quite well to
the true angle of rotation for $R<\frac{\pi}{2}$ ($\rho=0.64$). This
is unsurprising, because the closer the true angle is to zero, the
less the model has to rotate, and the less likely it will get stuck on
local maxima. Thus, it is more likely to locate the global maximum,
which generally corresponds to the true angle of rotation. For
$R>\frac{\pi}{2}$, we see an increasing tendency to under-rotate
($\rho=-0.42$); this is likely because it finds a local maxima and
ends prematurely.

@@

Similarly, because the nonparametric model actively attempts to obtain
an accurate estimate of $Z$, its estimate of $Z$ is indeed fairly
accurate ($\MSE{}=0.10$). This is slightly higher than the
nonparametric model, but is due to a few outliers. If we again exclude
stimuli for which $Z$ is greatly overestimated ($Z\geq 0.3$), the
nonparametric model's error decreases significantly ($\MSE{}=0.03$).

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/Z_accuracy.pdf}
  \caption{\textbf{Accuracy in estimating $Z$.} Each subplot shows the
    true (``gold standard'') value of $Z$ vs. the value estimated by
    the model. Black dotted lines indicate a perfect 1:1
    correspondence between the true and estimated values. All models
    perform well for $h_0$ stimuli, but exhibit different behavior for
    $h_1$ stimuli. The \naive{} model (left panel) tends to either
    greatly overestimate, or underestimate. The parametric model
    (center panel) is very accurate for about half the $h_1$ stimuli,
    and severely underestimates the rest. The nonparametric model
    (right panel) maintains a decent correspondence with the true $Z$,
    with the exception of a handful of outliers.}
  \label{fig:accuracy}
\end{figure}

@@

% \textbf{Acknowledgments}
% \TODO{}
